The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-05-08 paper.qmd
updated definition mistakes made section

@Hofman2020 performed an experiment where some participants asked to judge the effectiveness of a particular treatment when shown a sampling distribution of the mean or a prediction interval around the mean. A prediction interval indicates the uncertainty associated with treatment *for a particular person* and the sampling interval shown the uncertainty associated with the *average effectiveness of the treatment*. The authors reported that the group that was shown the sampling distribution judged the treatment as more effective than those that were shown a prediction interval, very likely because the prediction interval is always wider than the confidence interval. The importance of displaying the *most relevant* statistic and associated uncertainty is already well established in medicine, where doctors are required to understand a wide range of metrics that each communicate different risks to a patient [@Spiegelhalter2017]. Therefore, this work is not designed to commuicate to the field of medicine, but rather to the visualisation community who have historically struggled with the idea of uncertainty being related to inference. The perceived importance of this work by @Hofman2020 hinges on a widespread misunderstanding that the prediction and confidence interval *are* interchangeable despite performing inference on *very different* variables.

@Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty. The authors of this study were aware that uncertainty experiments were suffering from some kind of task dependence issue (due to other papers on the topic discussing it), but interpreted this as *context* dependence rather than dependence an inferential statistic. Participants were shown the six scenarios depicted in @fig-sketchy and asked to interpret what they believed the squiggly line indicates. A failing to understand the relationship between uncertainty and inference led the authors to add "sketchiness" to aspects of the visualization that could not possibly have an inferential statistic associated with it. It seems the authors did not consider the specific estimate that the uncertainty should represent, which is what led to visualisations that do not make conceptual sense. Participants rightfully assumed the sketchiness therefore represented something else, such alternative options or simply ignored it.

>>>>

The first common type of mistake is to present participants with insufficient information to answer a question due to the underlying belief that uncertainty is a latent product of the data. @Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty. Participants were shown the six scenarios depicted in @fig-sketchy and asked to interpret what they believed the squiggly line indicates. The authors were aware that uncertainty seemed to have some "task" dependence but interpreted this as *context* dependence rather than dependence on a particular inferential statistic. Therefore, "sketchiness" was added to aspects of the visualization that had no obvious inferential statistic associated with it, so participants simply ignored the "sketchiness" or assumed it represented something else, such as an alternative option. @Hofman2020 performed an experiment where participants were asked to judge the effectiveness of a particular treatment. One group was shown the prediction interval around the mean, which indicates the uncertainty associated with treatment *for a particular person* and another group was shown the the sampling interval which depicted the uncertainty associated with the *average effectiveness of the treatment*. There was no critical assessment by the authors on *why* these two distributions should be compared despite performing inference on completely different distributions. The contribution of this experiment hinges on the belief that sharing a common descriptive statistic (in this case, the mean) is grounds to compare two unrelated inferential statistics, a belief which is clearly false when viewing uncertainty through the lens of inference. Both studies highlight a clear belief that uncertainty is related to descriptive statistics, not inferential statistics.
